Estimating changes in temperature distributions in a large ensemble of climate simulations using quantile regression

### 2018-05-24 paper_revision.tex
Added Michaels comments

\abstract{
Understanding future changes in extreme temperature events in a transient climate is inherently challenging. A single model simulation is generally insufficient to empirically characterize the statistical properties of the underlying physical processes governing the climate. Ensembles of repeated simulations (with different initial conditions providing independence) not only alleviate this lack of data but also allow additional new approaches for characterizing changes in extremes. We present here one such new approach, using ensembles that allow characterizing changes in temperature distributions using a continuous representation of seasonality rather than breaking the dataset into seasonal blocks. That is, we assume that temperature distributions evolve smoothly both day-to-day over an annual cycle and year-to-year over longer secular trends. To demonstrate our method's utility, we analyze an ensemble of 50 simulations of the Community Earth System Model (CESM) under a scenario of increasing radiative forcing to 2100, focusing on North America. The results both confirm aspects of climate system behavior known from previous studies and also elucidate new features. Confirming results include that daily temperature bulk variability generally decreases  in wintertime in the continental mid- and high-latitudes ($> 40^{\circ}$). One new result is that these same wintertime distributions, the low tails ``stick'', i.e.\ experience a lesser reduction in variability, producing a more negative skew. Although the examples above concern temperature only, the technique is sufficiently general that it can be used to generate precise estimates of distribution changes in a broad range of climate variables by exploiting the power of ensembles.}

@@

Climate timeseries have generally been assumed to be separable into two components: randomness inherent in the underlying physical processes, which we call natural variability, and forced secular trends that follow from increasing concentrations of greenhouse gases.
Recently, the degree to which natural variability may itself be changing has received significant scientific interest \citep[e.g.][]{trenberth2011attribution, donat2012shifting,deser2012communication, thompson2015quantifying, kay2015community}. Potential changes in climate extremes, because of their heightened societal impacts, are of special concern \citep[e.g.][]{davison1990models, stott2004human, chavez2005generalized, eastoe2009modelling, otto2012reconciling, swain2014extraordinary, singh2014severe, trenberth2015attribution, diffenbaugh2015anthropogenic, huang2015changes, jalbert2017spatiotemporal}. 
However, detecting and understanding such changes on timescales relevant to human activities is inherently challenging and generally data-limited.
%identifying changes in climate variability related to human activities is inherently challenging. 
%The responses in the climate system are generally assumed to be separable into two components: 
The long equilibration time of the climate system means that on the timescales of interest to human society, the climate state will be evolving (or `transient'), so that its statistical properties are not stationary. Fully characterizing this evolving natural  variability of rare events is intrinsically challenging due to the limited amount of available observations or simulation data. Studies of future climate extremes often employ statistical extreme value theory to make inferences about rare events with modest amounts of data.
% XX some flipping of sentence order here, and made the key word "develop methods" instead of just "study" - EJM.

In this work, we study the entire distribution of temperatures in a transient climate, including rare events, by employing quantile regression on an ensemble of simulations of an identical forcing scenario from a single climate model.
% Each simulation has a different set of initial conditions to reflect the natural variability of the system. 
%Given sufficient care with choice of initial conditions, this approach can ensure that each simulation is statistically independent in terms of its natural variability.) 
Given sufficient care in choosing different initial conditions, such a large ensemble will reflect the natural variability of the system, since each simulation will be statistically independent in terms of its natural variability. 
The increased data provided by multiple simulations %XX allows instead of admits?
then admits more confident statements about changes in the statistical behavior of the system %XX compared to 
than can be made with a single simulation. While the use of ensembles is a relatively recent development, it is growing rapidly \citep[e.g.][]{deser2012uncertainty, deser2012communication,deser2014projecting,fischer2014detection, kay2015community,sriver2015effects, rodgers2015emergence, hagos2016projection}. \citet{deser2012uncertainty}, \citet{deser2012communication} and \citet{fischer2014detection} in particular discuss how ensembles help in distinguishing internal climate variability from anthropogenic effects and allow more accurate estimates for the forced model response.

% XX "Study" was used in two different ways in first two sentences. I changed to say "we will talk about 3 things" - EJM
%One under-exploited advantage of ensembles of multiple simulations is that the increased volume of data allows %XX study of examining the entire distribution of a climate variable.
Ensembles of multiple simulations offer at least three advantages that are currently under-exploited. The most obvious advantage is that the increased data volume allows examining the entire distribution of a climate variable.
 Studies of climate variability to date are generally divided between those that address the center of the distribution 
\citep[e.g.][]{semenov2002secular,raisanen2002co2, kitoh2009changes, screen2014arctic, schneider2015physics}, and those that address its tails \citep[e.g.][]{katz1992extreme, meehl2009relative, northrop2011threshold, davison2012statistical, huser2014space, trenberth2015attribution, huang2015estimating, jalbert2017spatiotemporal}, generally via extreme value theory. %In part the division has arisen from the data scarcity imposed by particular datasets.  %---- END CC
%High-resolution dynamically downscaled model studies \citep{XX,XX,XX} could in principle involve multiple runs, but their computational demands generally limit them to at most a few decades of present and future simulations. 
% XX Those studies 
A more limited body of studies address overall distributional changes in climate variables, but these generally focus on observations or observation-based data products, which are necessarily limited in terms of data amount and therefore require spatial or temporal aggregation \citep{donat2012shifting, stainforth2013mapping, chapman2013estimating, huybers2014us, mckinnon2016changing, rhines2017seasonally}. 
When studying model projections using ensembles, the large amount of data at each location allows us to accurately estimate changes in the distribution of climate variables (e.g. temperature) without spatial aggregation.

@@

\begin{figure} 
\includegraphics[width=\smallfigwidth]{schema.pdf}
\caption{\small{An illustration of concepts and values related to distributions used in this paper. The cartoon shows a positively skewed (or ``right-skewed'') probability distribution and the three quantile differences discussed in this paper, in the low tail, high tail, and middle of the distribution. 
The $p^{th}$ quantile in a distribution is the value such that the probability of being below it has probability $p$. Here $\Delta q_{\text{low}}$ is the difference between the $0.025^{th}$ and $0.05^{th}$ quantiles; the IQR or Interquartile Range that between the $0.25^{th}$ and $0.75^{th}$ quantiles; and $\Delta q_{\text{high}}$ that between the $0.95^{th}$ and $0.975^{th}$ quantiles. The values $\Delta q_{\text{low}}$ and $\Delta q_\text{high}$ quantify variability in extreme values while IQR quantifies variability in the bulk of the distribution.
}}            %
\label{schema}          %
\end{figure}

@@

The ensemble is especially appropriate for the purpose of studying variability because it is based on a ${\sim}10,000$ year pre-industrial control simulation. After a $\sim$4000 year spin-up using constant preindustrial conditions, we initialize 50 historical hindcasts (1850-2005) from snapshots of the coupled model state taken every 100 years, so that  the last hindcast is initialized after approximately 9000 years of the control simulation. We then
 extend each hindcast to 2100 using the Representative Concentration Pathway (RCP) 8.5 scenario. The 100-year gap between each new initialization ensures nearly independent ensemble members that fully capture internal variability within the coupled system. RCP8.5 corresponds to anthropogenic radiative forcing of roughly 8.5 W m$^{-2}$ by 2100 \citep{moss2010next}. More information about the model and ensemble design can be found in \citet{sriver2015effects}.

@@

% CC -- BEGIN edits below. "Validate" sounds a little more formal than what you're actually doing.
% CC -- if you say " With results for winter, each quantity shows a decent correspondence .." that implies that summer results DONT show a correspondence. 
To evaluate whether the CESM simulations provide sufficiently realistic temperature distributions for the purpose of this analysis, we compare CESM temperatures with those from the ERA-Interim (European Reanalysis) data product \citep{dee2011era}. Figure \ref{model_reanalysis_world} shows the model/reanalysis comparison for winter; for summer see Supplementary Online Material Figure \ref{model_reanalysis_world_JJA}. The model underestimates variability in some places, and produces excessively cold winter temperatures in the Arctic. The resulting exaggerated temperature gradients contribute to excess variability and excessively negative skew in the northern mid-latitudes.
(Skewness is proportional to the cube of temperature after subtracting off the average seasonal temperature; see Appendix \ref{app:moments}.) %% ejm changed from cubes to cube.
Throughout this work, we will show in-depth analysis from three locations with distinct temperature distributions to highlight our proposed method (\textbf{a}, \textbf{b}, and \textbf{c} shown in Figure \ref{model_reanalysis_world}). See Supplementary Online Material Figure \ref{rea_temp_locations} for comparison of model and reanalysis temperature distributions in both summer and winter for these locations. 
% CC could add a sentence saying this is still OK for your purpose if you want, but I think you can skip that

@@

% CC BEGIN - can't say "annual change", it's confusing. 
In the methodology presented here, we 
model temperature at each location as a function of both seasonality and long-term change of the annual mean climate. %where the latter captures the transient nature of the climate. 
We use two independent variables, with seasonality represented by a variable $d$, the day of the year (spanning values $1$ to $365$), and change in annual mean temperature represented by a variable $t$, years elapsed since $1850$ (spanning $0$ to $250$ for these scenarios).
We thus assume that each temperature quantile can be described by two sets of basis functions that represent the two variables' independent relationships with temperature (called here $\{f_i(d)\}$ and $\{g_j(t)\}$), and interaction terms $h_i(d)s_j(t)$, where 
$f_i, g_j, h_i,$ and $s_j$ are all smooth functions of the appropriate variable. % CC I would say 'assume'; if you say 'create' you have to discuss functional form here. And best to avoid talking about i and j here since you do it below... can you hold off?
%The span of $i$ and $j$ is determined by how many basis functions are appropriate for the model (discussed below).
The interaction terms are required to capture effects in which long-term temperature evolution differs between seasons, e.g.\
%Allowing for interaction between seasonality and long term change allows capturing effects such as 
the robust projection that winter temperatures warm more than summer temperatures. 
To impose our smoothness condition, we assume that $f_i, g_j, h_i,$ and $s_j$ are piecewise cubic polynomials with a continuous second derivative, also called splines. \citep[For a review of cubic polynomial basis functions, see][Chapter 5.]{ESL}
Because the seasonality variable $d$ is periodic, its basis functions are also assumed periodic. 
For more details, see Appendix \ref{app:model}a.

We choose the number of basis functions by evaluating a metric representing model sufficiency. Our model sufficiency criterion is aimed at capturing the long term underlying signal. We do not require estimated quantile functions to capture transient events during the historical period like volcanic eruptions. Details on how we select the number of basis functions is given in Appendix \ref{app:model}b. In our climate simulation output, the intra-seasonal effect requires more detailed modeling than the inter-seasonal effect. In the results shown here, we fit the model with 15 terms (that is, basis functions) for the main seasonal effect $\{f_i\}$, but the interaction terms require less seasonal complexity, so we use only 3 terms for $\{h_i\}$.  
We use 4 terms for both the temporal change $\{g_j\}$ and the interaction terms $\{s_j\}$. 
That is, modeling long-term change generally requires fewer terms than modeling seasonality. In summary, we use 32 basis functions in total including an intercept term. We then fit each $q^{th}$ quantile of temperature % --- END CC
\begin{equation}
T_q(d, t) = \alpha + \sum_ia_if_i(d) + \sum_jb_j g_j(t) + \sum_{i,j}c_{i,j} h_i(d)s_j(t),
\end{equation}
 where all of the coefficients depend on $q$ but we suppress the dependence for convenience. This fit determines coefficients $a_i$, $b_j$, $c_{i,j}$ for each quantile at each location.

 To simplify notation, we construct a matrix $X$ where each column contains a basis function and each row refers to a unique value of $d$ and $t$. Using this matrix, $X$, we construct our temperature model in vectorized form,
 \begin{equation}
 \label{eq:matrix_version}
 T_q = X\beta_q, 
 \end{equation}
 where $\beta_q$ contains the basis coefficients $a_i, b_j, c_{i,j}$. If the total number of basis functions is $p$ and the total number of observations is $n$, then the predictor matrix $X$ will have dimensions $n \times p$ and $\beta_q$ will be a $p$-length vector. To get a confidence interval around $T_q$, we re-estimate the coefficients, $\beta_q$, using a resampled data set. Because we have 50 simulations we resample the data by drawing whole simulations from our ensemble of 50 simulations. Each simulation is treated as a contiguous block of data, and the dependency structure within these blocks is maintained when resampling whole simulations at a time. By repeating this resampling and re-estimation procedure 100 times we obtain pointwise confidence bands around each estimated $T_q$. Appendix \ref{app:model}c provides further details about uncertainty quantification.

@@

\begin{figure}[t]       % it is a macro to save typing later
\centerline{\includegraphics[width=\smallfigwidth]{Estimate_3loc.pdf}}
\caption{\small{Illustration of results of our quantile estimation procedure using the 50-member CESM ensemble. The figure shows ensemble daily mean temperatures for the year 1850 for the three representative locations \textbf{a}, \textbf{b}, and \textbf{c} defined in Figure \ref{model_reanalysis_world}. The ensemble provides 50 points per day but for clarity we show only $10\%$ of the data. Solid lines show the median daily temperature and dashed lines the $2.5^{th}$ and $97.5^{th}$ quantiles estimated by our procedure. %Our procedure captures smoothly-varying seasonal cycles in variability at each location, with 
(Note the higher variabilities in winter.)  Note that the location of the points exceeding the smooth median estimate are approximately uniform across time (notwithstanding the amplitudes of residuals), suggesting that the quantile estimate is accurate for each day. XX put in Michael's text here
At all sites, the estimated quantile curves capture the seasonally changing patterns in the distributions reasonably well.
}}

@@

\begin{figure}[ht]
\centerline{\includegraphics[width=\smallfigwidth]{meanStdSkew_Winter_wcrosses.pdf}}
\caption{\small{Initial temperature distribution properties (left) and their changes over time (right) in the CESM ensemble RCP8.5 model runs, for aggregate wintertime (DJF) daily temperature. Initial (``pre-industrial'') and final periods are defined as in Figure 4, as 15-year periods 1850--1864 and 2086--2100.  Distributional moments (mean, standard deviation, and skewness) are defined as in Figure 2. Units on the top two rows are degrees Celsius, while the bottom row showing skewness is dimensionless. Gray crosses mark locations where the changes are not significant at the $0.05$ level, obtained by resampling the set of $50$ simulations (with replacement) and recalculating the sample moments. \emph{Top right}: Mean temperature universally increases; note that color scale begins at XX degrees. Extreme warming in the Hudson's Bay region occurs where the model is biased low in present-day simulations. \emph{Middle right}: As expected, standard deviation decreases strongly at higher latitudes. \emph{Bottom right}: Changes in winter skewness show a dipole pattern, which enhances negative skew above $\sim 40 ^\circ$ but reduces it at lower latitudes. }}
\label{meanStdSkew_Winter}
\end{figure}

@@

With a smooth estimate of quantiles of average temperature, we show that the onset of spring, as measured by the first day of the year where average temperature reaches $-2.2^{\circ}$C \citep{pearse2017statistical}, occurs earlier in the year as the climate warms in the Detroit area (see Figure \ref{fig:FirstDayAbove2_2}). The lower quantiles seem to be progessing faster than the median quantiles, with the $25^{th}$ quantile hitting the $-2.2^{\circ}$C mark at a rate of approximately 15 days earlier per decade at present times. Note also that the $50^{th}$ quantile never goes below the threshold after year 2080. It is unclear how to produce the equivalent results using existing methods of segmenting average temperature into seasons. For instance, if we were to look at quantiles of average temperature during winter the edges of the season would pull the overall quantile estimates up and prematurely estimate the onset of spring.

\begin{figure}[ht]
\centerline{\includegraphics[width=\smallfigwidth]{FirstDayAbove2_2.pdf}}
\caption{\small{First day above $-2.2^\circ$C for each year from 1850-2100 as measured by fitting quantiles average temperature of the CESM ensemble data set. Three quantiles are shown to capture the spread of the distribution, $.5$ (green), $.25$ (red) and $.05$ (black).}}
\label{fig:FirstDayAbove2_2}
\end{figure}

%To what extent the distributional characteristics are changing, and whether these changes are significant. 
Our methodology for quantile estimation provides additional information that helps to
 quantify how temperature distributions are changing and to estimate the uncertainty associated with each change. 
%The goal is to understand how the temperature distribution is changing, both in the middle and in the extremes (or tails) of the distribution. 
We can evaluate not only bulk variability  -- the interquartile range (IQR), the difference between the 0.25th and 0.75th quantiles -- but differences between any two quantiles. 
We therefore evaluate the difference between two low or high quantiles, denoted $\Delta q_{low}$ and $\Delta q_{high}$, which measure tail variability in the same way that interquartile range measures the variability of the bulk distribution.
If the skewness of a distribution changes over time, then future distributions are not simply scaled versions of present distributions. That is, their tail variabilities must change differently than does the IQR.
  In the case of the northern mid-latitudes winter temperatures shown in Figure \ref{meanStdSkew_Winter}, where distributions become more negatively skewed as bulk variability decreases in the future, the effect could result from either/both a low tail contracting less than the bulk (or actually increasing), or a high tail contracting more than the bulk. Our methodology allows readily differentiating these cases.

To assess whether the high tail and/or the low tail is driving changes in skewness, we consider the fractional changes in low, high, and bulk variability.
%Bulk variability, e.g. the interquartile range (IQR), is commonly linked to the tail variability. For example, a decrease in the bulk variability as a result of an overall scale change will produce a corresponding decrease in the tail variability. However, if for example the tail contracts less the bulk, there will be a skewness change. To show how variability at the different quantiles change with respect to the initial value, 
If we denote the initial and final quantile difference as $\Delta q_{p, i}$ and $\Delta q_{p, f}$ at the $p^{th}$ percentile, the temporal change in quantile differences relative to the initial year is then
\begin{equation}
	\rho = \frac{\Delta q_{p, f} - \Delta q_{p, i}}{\Delta q_{p, i}}.
\label{eq:relnorm}
\end{equation}
 Because we model the complete temperature distribution for each day of the year for all years, we choose a representative day to understand winter and summer changes (Jan 1 and July 5, respectively), and consider the difference between the beginning and end of the scenarios, the years 1850 and 2000. For these representative days, we show in Figure \ref{winter_summer_var_avg} the fractional variability changes of $\rho$ for low and high tails as well as the IQR.

>>>>

\abstract{
Understanding future changes in extreme temperature events in a transient climate is inherently challenging. A single model simulation is generally insufficient to empirically characterize the statistical properties of the underlying physical processes governing the climate. Ensembles of repeated simulations (with different initial conditions providing independence) greatly expand the amount of data available, which in turn allows new approaches for characterizing changes in extremes. We present here one such new approach, using ensembles that allow characterizing changes in temperature distributions using a continuous representation of seasonality rather than breaking the dataset into seasonal blocks. That is, we assume that temperature distributions evolve smoothly both day-to-day over an annual cycle and year-to-year over longer secular trends. To demonstrate our method's utility, we analyze an ensemble of 50 simulations of the Community Earth System Model (CESM) under a scenario of increasing radiative forcing to 2100, focusing on North America. The results both confirm aspects of climate system behavior known from previous studies and also elucidate new features. Confirming results include that daily temperature bulk variability generally decreases in wintertime in the continental mid- and high-latitudes ($> 40^{\circ}$). One new result is that these same wintertime distributions, the low tails ``stick'', i.e.\ experience a lesser reduction in variability, producing a more negative skew. Although the examples above concern temperature only, the technique is sufficiently general that it can be used to generate precise estimates of distribution changes in a broad range of climate variables by exploiting the power of ensembles.}

@@

Climate time series have generally been assumed to be separable into two components: randomness inherent in the underlying physical processes, which we call natural variability, and climatic trends, e.g. in the form of forced secular trends that follow from increasing concentrations of greenhouse gases or seasonal trends.
Recently, the degree to which natural variability may itself be changing has received significant scientific interest \citep[e.g.][]{trenberth2011attribution, donat2012shifting,deser2012communication, thompson2015quantifying, kay2015community}. Potential changes in climate extremes, because of their heightened societal impacts, are of special concern \citep[e.g.][]{davison1990models, stott2004human, chavez2005generalized, eastoe2009modelling, otto2012reconciling, swain2014extraordinary, singh2014severe, trenberth2015attribution, diffenbaugh2015anthropogenic, huang2015changes, jalbert2017spatiotemporal}. 
However, fully characterizing this evolving natural variability of rare events is intrinsically challenging due to the limited amount of available observations or simulation data.
%identifying changes in climate variability related to human activities is inherently challenging. 
The long equilibration time of the climate system means that on the timescales of interest to human society, the climate state will be evolving, so that its statistical properties are not stationary. Studies of future climate extremes often employ statistical extreme value theory to make inferences about rare events with modest amounts of data \citep{swain2014extraordinary}.

In this work, we study the entire distribution of temperatures in a transient climate, including rare events, by employing quantile regression on an ensemble of simulations of an identical forcing scenario from a single climate model. Given sufficient care in choosing different initial conditions, such a large ensemble will reflect the natural variability of the system, since each simulation will be statistically independent in terms of its natural variability. 
The increased data provided by multiple simulations then admits more confident statements about changes in the statistical behavior of the system than can be made with a single simulation. While the use of ensembles is a relatively recent development, it is growing rapidly \citep[e.g.][]{deser2012uncertainty, deser2012communication,deser2014projecting,fischer2014detection, kay2015community,sriver2015effects, rodgers2015emergence, hagos2016projection}. \citet{deser2012uncertainty}, \citet{deser2012communication} and \citet{fischer2014detection} in particular discuss how ensembles help in distinguishing internal climate variability from anthropogenic effects and allow more accurate estimates for the forced model response.

% XX "Study" was used in two different ways in first two sentences. I changed to say "we will talk about 3 things" - EJM
%One under-exploited advantage of ensembles of multiple simulations is that the increased volume of data allows %XX study of examining the entire distribution of a climate variable.
Ensembles of multiple simulations offer at least three advantages that are currently under-exploited. The most obvious advantage is that the increased data volume allows examining the entire distribution of a climate variable.
 Studies of climate variability to date are generally divided between those that address the center of the distribution 
\citep[e.g.][]{semenov2002secular,raisanen2002co2, kitoh2009changes, screen2014arctic, schneider2015physics}, and those that address its tails \citep[e.g.][]{katz1992extreme, meehl2009relative, northrop2011threshold, davison2012statistical, huser2014space, trenberth2015attribution, huang2015estimating, jalbert2017spatiotemporal}, generally via extreme value theory. %In part the division has arisen from the data scarcity imposed by particular datasets.  %---- END CC
%High-resolution dynamically downscaled model studies \citep{XX,XX,XX} could in principle involve multiple runs, but their computational demands generally limit them to at most a few decades of present and future simulations. 
% XX Those studies 
A more limited body of studies address overall distributional changes in climate variables, but these generally focus on observations or observation-based data products, which are necessarily limited in terms of data amount and therefore require spatial or temporal aggregation \citep{donat2012shifting, stainforth2013mapping, chapman2013estimating, huybers2014us, mckinnon2016changing, rhines2017seasonally}. Aggregating data spatio-temporally requires stationarity assumptions of the signal or explicitly modeling the spatio-temporal dependence. When studying model projections using ensembles, the large amount of data at each location allows us to accurately estimate changes in the distribution of climate variables (e.g. temperature) without spatial aggregation.

@@

\begin{figure} 
\includegraphics[width=\smallfigwidth]{schema.pdf}
\caption{\small{An illustration of concepts and values related to distributions used in this paper. The cartoon shows a positively skewed (or ``right-skewed'') probability distribution and the three quantile differences discussed in this paper, in the low tail, high tail, and middle of the distribution. 
The $q$ quantile in a distribution is the value such that the probability of being below it has probability $q$. Here $\Delta q_{\text{low}}$ is the difference between the $0.025$ and $0.05$ quantiles; the IQR or Interquartile Range that between the $0.25$ and $0.75$ quantiles; and $\Delta q_{\text{high}}$ that between the $0.95$ and $0.975$ quantiles. The values $\Delta q_{\text{low}}$ and $\Delta q_\text{high}$ quantify variability in extreme values while IQR quantifies variability in the bulk of the distribution.
}}            %
\label{schema}          %
\end{figure}

@@

The ensemble is especially appropriate for the purpose of studying variability because it is based on a ${\sim}10,000$ year pre-industrial control simulation. After a $\sim$4000 year spin-up using constant preindustrial conditions, 50 historical hindcasts (1850-2005) are initialized from snapshots of the coupled model state taken every 100 years, so that the last hindcast is initialized after approximately 9000 years of the control simulation. Each hindcast is then extended to 2100 using the Representative Concentration Pathway (RCP) 8.5 scenario. The 100-year gap between each new initialization ensures nearly independent ensemble members that fully capture internal variability within the coupled system. RCP8.5 corresponds to anthropogenic radiative forcing of roughly 8.5 W m$^{-2}$ by 2100 \citep{moss2010next}. More information about the model and ensemble design can be found in \citet{sriver2015effects}.

@@

% CC -- BEGIN edits below. "Validate" sounds a little more formal than what you're actually doing.
% CC -- if you say " With results for winter, each quantity shows a decent correspondence .." that implies that summer results DONT show a correspondence. 
To evaluate whether the CESM simulations provide sufficiently realistic temperature distributions for the purpose of this analysis, we compare CESM temperatures with those from the ERA-Interim (European Reanalysis) data product \citep{dee2011era}. Figure \ref{model_reanalysis_world} shows the model/reanalysis comparison for winter; for summer see Supplementary Online Material Figure \ref{model_reanalysis_world_JJA}. The model underestimates variability in some places, and produces excessively cold winter temperatures in the Arctic. The resulting exaggerated temperature gradients contribute to excess variability and excessively negative skew in the northern mid-latitudes.
Skewness is proportional to the cube of temperature after subtracting off the average seasonal temperature; see Appendix \ref{app:moments}. %% ejm changed from cubes to cube.
Throughout this work, we will show in-depth analysis from three locations with distinct temperature distributions to highlight our proposed method (\textbf{a}, \textbf{b}, and \textbf{c} shown in Figure \ref{model_reanalysis_world}). See Supplementary Online Material Figure \ref{rea_temp_locations} for comparison of model and reanalysis temperature distributions in both summer and winter for these locations. 
% CC could add a sentence saying this is still OK for your purpose if you want, but I think you can skip that

@@

% CC BEGIN - can't say "annual change", it's confusing. 
In the methodology presented here, we model temperature at each location as a function of both seasonality and long-term change of the annual temperature distribution. %where the latter captures the transient nature of the climate. 
We use two independent variables, with seasonality represented by a variable $d$, the day of the year (spanning values $1$ to $365$), and change in annual temperature represented by a variable $t$, years elapsed since $1850$ (spanning $0$ to $250$ for these scenarios).
We thus assume that each temperature quantile can be described by two sets of basis functions that represent the two variables' independent relationships with temperature (called here $\{f_i(d)\}$ and $\{g_j(t)\}$), and interaction terms $h_i(d)s_j(t)$, where 
$f_i, g_j, h_i,$ and $s_j$ are all smooth functions of the appropriate variable. % CC I would say 'assume'; if you say 'create' you have to discuss functional form here. And best to avoid talking about i and j here since you do it below... can you hold off?
%The span of $i$ and $j$ is determined by how many basis functions are appropriate for the model (discussed below).
The interaction terms are required to capture effects in which long-term temperature evolution differs between seasons, e.g.\
%Allowing for interaction between seasonality and long term change allows capturing effects such as 
the robust projection that winter temperatures warm more than summer temperatures. 
To impose our smoothness condition, we assume that $f_i, g_j, h_i,$ and $s_j$ are piecewise cubic polynomials with a continuous second derivative, also called splines. \citep[For a review of cubic polynomial basis functions, see][Chapter 5.]{ESL}
Because the seasonality variable $d$ is periodic, its basis functions are also assumed periodic. 
For more details, see Appendix \ref{app:model}a.

We choose the number of basis functions by evaluating a metric representing model adequacy. Our model sufficiency criterion is aimed at capturing the long term underlying signal. We do not require estimated quantile functions to capture transient events during the historical period like volcanic eruptions. Details on how we select the number of basis functions is given in Appendix \ref{app:model}b. In our climate simulation output, the intra-seasonal effect requires more detailed modeling than the inter-seasonal effect. In the results shown here, we fit the model with 15 terms (that is, basis functions) for the main seasonal effect $\{f_i\}$, but the interaction terms require less seasonal complexity, so we use only 3 terms for $\{h_i\}$.  
We use 4 terms for both the temporal change $\{g_j\}$ and the interaction terms $\{s_j\}$. 
That is, modeling long-term change generally requires fewer terms than modeling seasonality. In summary, we use 32 basis functions in total including an intercept term. We then fit each $q$ quantile of temperature % --- END CC
\begin{equation}
T_q(d, t) = \alpha + \sum_ia_if_i(d) + \sum_jb_j g_j(t) + \sum_{i,j}c_{i,j} h_i(d)s_j(t),
\end{equation}
 where all of the coefficients depend on $q$ but we suppress the dependence for convenience. This fit determines coefficients $a_i$, $b_j$, $c_{i,j}$ for each quantile at each location.

 To simplify notation, we construct a matrix $X$ where each column contains a basis function and each row refers to a unique value of $d$ and $t$. Using this matrix, $X$, we construct our temperature model in vectorized form,
 \begin{equation}
 \label{eq:matrix_version}
 T_q = X\beta_q, 
 \end{equation}
 where $\beta_q$ contains the basis coefficients $a_i, b_j, c_{i,j}$. The predictor matrix $X$ will have $32$ columns, each corresponding to one basis function, and $365 \times 250 \times 50$ rows, where each row is a daily average temperature. Similarly, $\beta_q$ will be a $32$-length vector. To get a confidence interval around $T_q$, we re-estimate the coefficients, $\beta_q$, using a resampled data set. Because we have 50 simulations we resample the data by drawing whole simulations from our ensemble of $50$ simulations. By resampling complete realizations, the dependency structure within realizations is maintained in the resampled data. Thus, repeating this resampling and re-estimation procedure 100 times yields pointwise confidence bands around each estimated $T_q$. Appendix \ref{app:model}c provides further details about uncertainty quantification.

@@

\begin{figure}[t]       % it is a macro to save typing later
\centerline{\includegraphics[width=\smallfigwidth]{Estimate_3loc.pdf}}
\caption{\small{Illustration of results of our quantile estimation procedure using the 50-member CESM ensemble. The figure shows ensemble daily mean temperatures for the year 1850 for the three representative locations \textbf{a}, \textbf{b}, and \textbf{c} defined in Figure \ref{model_reanalysis_world}. The ensemble provides 50 points per day but for clarity we show only $10\%$ of the data. Solid lines show the median daily temperature and dashed lines the $.025$ and $.975$ quantiles estimated by our procedure. %Our procedure captures smoothly-varying seasonal cycles in variability at each location, with 
(Note the higher variabilities in winter.)  Note that the location of the points exceeding the smooth median estimate are approximately uniform across time (notwithstanding the amplitudes of residuals), suggesting that the quantile estimate is accurate for each day. XX put in Michael's text here
At all sites, the estimated quantile curves capture the seasonally changing patterns in the distributions reasonably well.
}}

@@

\begin{figure}[ht]
\centerline{\includegraphics[width=\smallfigwidth]{meanStdSkew_Winter_wcrosses.pdf}}
\caption{\small{Initial temperature distribution properties (left) and their changes over time (right) in the CESM ensemble RCP8.5 model runs, for aggregate wintertime (DJF) daily temperature. Initial (``pre-industrial'') and final periods are defined as in Figure 4, as 15-year periods 1850--1864 and 2086--2100.  Distributional moments (mean, standard deviation, and skewness) are defined as in Figure 2. Units on the top two rows are degrees Celsius, while the bottom row showing skewness is dimensionless. Gray crosses mark locations where the changes are not significant at the $0.05$ level, obtained by resampling the set of $50$ simulations (with replacement) and recalculating the sample moments. \emph{Top right}: Mean temperature universally increases. Extreme warming in the Hudson's Bay region occurs where the model is biased low in present-day simulations. \emph{Middle right}: As expected, standard deviation decreases strongly at higher latitudes. \emph{Bottom right}: Changes in winter skewness show a dipole pattern, which enhances negative skew above $\sim 40 ^\circ$ but reduces it at lower latitudes. }}
\label{meanStdSkew_Winter}
\end{figure}

@@

With a smooth estimate of quantiles of average temperature, we show that the onset of spring, as measured by the first day of the year where the $.5$ quantile estimate reaches $-2.2^{\circ}$C \citep{pearse2017statistical}, occurs earlier in the year as the climate warms in the Detroit area (see Figure \ref{fig:FirstDayAboveFreezing}). The lower quantiles seem to be progessing faster than the $.5$ quantile, with the $.25$ quantile hitting the $-2.2^{\circ}$C mark at a rate of approximately 15 days earlier per decade at present times. Note also that the $.5$ quantile never goes below the threshold after year 2080. It is unclear how to produce the equivalent results using existing methods of segmenting average temperature into seasons. For instance, if we were to look at quantiles of average temperature during winter the edges of the season would pull the overall quantile estimates up and prematurely estimate the onset of spring. Moreover getting information regarding the average temperature of a specific day would be impossible with seasonal averaging.

\begin{figure}[ht]
\centerline{\includegraphics[width=\smallfigwidth]{FirstDayAboveFreezing.pdf}}
\caption{\small{First day above freezing (solid lines) and $-2.2^\circ$C (dashed lines) for each year from 1850-2100 as measured by fitting quantiles to average daily temperature of the CESM ensemble data set. Three quantiles are shown to capture the spread of the distribution, $.5$ (green), $.25$ (red) and $.05$ (black).}}
\label{fig:FirstDayAboveFreezing}
\end{figure}

%To what extent the distributional characteristics are changing, and whether these changes are significant. 
Our methodology for quantile estimation provides additional information that helps to
 quantify how temperature distributions are changing and to estimate the uncertainty associated with each change. 
%The goal is to understand how the temperature distribution is changing, both in the middle and in the extremes (or tails) of the distribution. 
We can evaluate not only bulk variability  -- the interquartile range (IQR), the difference between the $0.25$ and $0.75$ quantiles -- but differences between any two quantiles. 
We therefore evaluate the difference between two low or high quantiles, denoted $\Delta q_{low}$ and $\Delta q_{high}$, which measure tail variability in the same way that interquartile range measures the variability of the bulk distribution.
If the skewness of a distribution changes over time, then future distributions are not simply scaled versions of present distributions. That is, their tail variabilities must change differently than does the IQR.
  In the case of the northern mid-latitudes winter temperatures shown in Figure \ref{meanStdSkew_Winter}, where distributions become more negatively skewed as bulk variability decreases in the future, the effect could result from either/both a low tail contracting less than the bulk (or actually increasing), or a high tail contracting more than the bulk. Our methodology allows readily differentiating these cases.

To assess whether the high tail and/or the low tail is driving changes in skewness, we consider the fractional changes in low, high, and bulk variability.
%Bulk variability, e.g. the interquartile range (IQR), is commonly linked to the tail variability. For example, a decrease in the bulk variability as a result of an overall scale change will produce a corresponding decrease in the tail variability. However, if for example the tail contracts less the bulk, there will be a skewness change. To show how variability at the different quantiles change with respect to the initial value, 
If we denote the initial and final quantile difference as $\Delta q_{q, i}$ and $\Delta q_{q, f}$ at the $q$ quantile, the temporal change in quantile differences relative to the initial year is then
\begin{equation}
	\rho = \frac{\Delta q_{q, f} - \Delta q_{q, i}}{\Delta q_{q, i}}.
\label{eq:relnorm}
\end{equation}
 Because we model the complete temperature distribution for each day of the year for all years, we choose a representative day to understand winter and summer changes (Jan 1 and July 5, respectively), and consider the difference between the beginning and end of the scenarios, the years 1850 and 2000. For these representative days, we show in Figure \ref{winter_summer_var_avg} the fractional variability changes of $\rho$ for low and high tails as well as the IQR.
